---
title: "2 Biological Sequences"
author: "Olivia Rippee"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Biostrings)
library(BSgenome.Hsapiens.UCSC.hg19)
library(AnnotationHub)
library(GenomicRanges)
library(GenomicFeatures)
library(rtracklayer)
library(IRanges)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)

# Human hg19 reference genome
Hsapiens <- BSgenome.Hsapiens.UCSC.hg19

# AnnotationHub
ah <- AnnotationHub()

```

# What is the GC content of “chr22” in the “hg19” build of the human genome?
## The reference genome includes “N” bases; exclude those from the calculation.

```{r}

# Count A, C, G, and T bases
total_bases <- sum(letterFrequency(Hsapiens$chr22, c("A", "C", "G", "T")))

# Count G and C bases
gc_bases <- sum(letterFrequency(Hsapiens$chr22, c("G", "C")))

# Calculate GC content
(gc_content <- gc_bases / total_bases) # 0.4798807

```

# What is the mean GC content of H3K27me3 “narrowPeak” regions from Epigenomics Roadmap from the H1 stem cell line on chr 22?
## Compute the GC content for each peak region as a percentage and then average those percentages to obtain a number between 0 and 1.

```{r}

# Find the H1/E003 H3K27me3 narrowPeak dataset
H3K27me3_qh <- query(ah, c("H3K27me3", "E003", "narrowPeak"))

# Retrieve the narrowPeak GRanges object
H3K27me3_data <- H3K27me3_qh[[1]]

# Keep only chr22 peaks
H3K27me3_data_chr22 <- H3K27me3_data[seqnames(H3K27me3_data) == "chr22"]

# Extract the corresponding DNA sequences
H3K27me3_data_chr22_seq <- getSeq(Hsapiens, H3K27me3_data_chr22)

# Calculate GC content for each peak
gcContents <- letterFrequency(H3K27me3_data_chr22_seq, "GC", as.prob = TRUE)

# Average GC content across peaks
(meanGcContents <- mean(gcContents)) # 0.528866

```

# What is the correlation between GC content and “signalValue” of these regions (on chr22)?

```{r}

# Extract signalValue from the narrowPeak metadata.
sigV <- mcols(H3K27me3_data_chr22_seq)$signalValue

# Calculate Pearson correlation.
(gc_signal_cor <- cor(sigV, gcContents, use = "complete.obs")) # 0.004467924

```

# What is the correlation between the “signalValue” of the “narrowPeak” regions and the average “fc.signal” across the same regions?
## First compute the average fc.signal across each peak using Views,  then correlate those averages with signalValue

```{r}

# Find the E003 H3K27me3 fc.signal BigWig file
H3K27me3_fc <- query(ah, c("H3K27me3", "E003", "fc.signal"))

# Retrieve the BigWigFile
H3K27me3_fc_data <- H3K27me3_fc[[1]]

# Define chr22
gr_chr22 <- GRanges(seqnames = "chr22",
  ranges = IRanges(start = 1, end = seqlengths(Hsapiens)["chr22"]))

# Import the fc.signal track for chr22 as an Rle
H3K27me3_fc_rel <- import(H3K27me3_fc_data, which = gr_chr22, as = "Rle")

# Extract the chr22 signal
H3K27me3_fc_rel_chr22 <- H3K27me3_fc_rel$chr22

# Create Views corresponding to each narrowPeak region
fc.signal_chr22 <- Views(
  H3K27me3_fc_rel_chr22,
  start = start(H3K27me3_data_chr22),
  end = end(H3K27me3_data_chr22))

# Calculate the mean fc.signal for each peak
fc.signal_mean <- viewMeans(fc.signal_chr22)

# Correlate mean fc.signal with narrowPeak signalValue
(signal_fc_cor <- cor(fc.signal_mean, sigV, use = "complete.obs")) # 0.9149614

```

# How many bases on chr22 have an fc.signal greater than or equal to 1?

```{r}

(bases_fc_ge_1 <- sum(H3K27me3_fc_rel_chr22 >= 1)) # 10914671

```

# Identify the regions of the genome where the signal in E003 is 0.5 or lower and the signal in E055 is 2 or higher.

```{r}

# These are regions that gain H3K27me3 signal upon differentiation

# Find the E055 H3K27me3 fc.signal dataset
H3K27me3_fc_E055 <- query(ah, c("H3K27me3", "E055", "fc.signal"))

# Find the record whose title is the E055 H3K27me3 fc.signal BigWig
e055_index <- which(
  grepl("E055-H3K27me3.fc.signal.bigwig",
    mcols(H3K27me3_fc_E055)$title))[1]

H3K27me3_fc_E055_data <- H3K27me3_fc_E055[[e055_index]]

# Import E055 signal on chr22
H3K27me3_fc_rel_E055 <- import(H3K27me3_fc_E055_data, which = gr_chr22, as = "Rle")

H3K27me3_fc_rel_E055_chr22 <- H3K27me3_fc_rel_E055$chr22

# Identify regions where E003 signal <= 0.5
region_E003 <- slice(H3K27me3_fc_rel_chr22, upper = 0.5)

# Identify regions where E055 signal >= 2
region_E055 <- slice(H3K27me3_fc_rel_E055_chr22, lower = 2)

# Convert Views to IRanges before intersecting
region_E003 <- as(region_E003, "IRanges")
region_E055 <- as(region_E055, "IRanges")

# Find regions satisfying both conditions
inter_region <- intersect(region_E003, region_E055)

# Count the number of bases in those regions
(bases_gain_H3K27me3 <- sum(width(inter_region))) # 1869937

```

# What is the average observed-to-expected ratio of CpG dinucleotides for CpG Islands on chromosome 22?

```{r}

# Find the hg19 CpG Island dataset.
CpGIsland <- query(ah, c("hg19", "CpG Islands"))

# Retrieve the CpG Island GRanges object
CpGIsland_data <- CpGIsland[[1]]

# Keep only CpG Islands on chr22
CpGIsland_data_chr22 <- CpGIsland_data[seqnames(CpGIsland_data) == "chr22"]

# Extract DNA sequences
CpGIsland_data_chr22_seq <- getSeq(Hsapiens, CpGIsland_data_chr22)

# Length of each CpG Island
region_length <- width(CpGIsland_data_chr22_seq)

# Observed frequency of CG dinucleotides.
# "CG" is the 7th dinucleotide in dinucleotideFrequency().
observed_CG <- dinucleotideFrequency(CpGIsland_data_chr22_seq)[, "CG"]

observed_CG_frequency <- observed_CG / region_length

# Frequency of C and G
freq_C <- letterFrequency(CpGIsland_data_chr22_seq, "C")
freq_G <- letterFrequency(CpGIsland_data_chr22_seq, "G")

# Expected CpG frequency = frequency(C) * frequency(G)
expected_CG_frequency <- ((freq_C / region_length) * (freq_G / region_length))

# Observed-to-expected ratio for each island.
observed_expected_ratio <- (observed_CG_frequency / expected_CG_frequency)

# Average across all CpG Islands.
(mean_observed_expected <- mean(observed_expected_ratio, na.rm = TRUE)) # 0.8340929

```

# How many TATA boxes are there on chr 22 of build hg19 of the human genome?
## TATA here is "TATAAA". Search both the forward and reverse strands.

```{r}

# Count TATAAA on the forward strand
tata_forward <- countPattern("TATAAA", Hsapiens$chr22)

# Count TATAAA on the reverse-complement strand
tata_reverse <- countPattern("TATAAA", reverseComplement(Hsapiens$chr22))

# Total TATAAA sites on either strand
(tata_boxes <- tata_forward + tata_reverse) # 27263

```

# How many promoters of transcripts on chromosome 22 containing a coding sequence contain a TATA box on the same strand as the transcript?
## Promoter: 900 bp upstream and 100 bp downstream of the transcription start site

```{r}

txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene

# Get transcripts on chr22
tx_chr22 <- transcripts(txdb, filter = list(tx_chrom = "chr22"))

# Determine which transcripts contain coding sequence
tx_lengths <- transcriptLengths(txdb, with.cds_len = TRUE)

coding_tx_ids <- tx_lengths$tx_id[tx_lengths$cds_len > 0]

# Keep only coding transcripts on chr22
coding_tx_chr22 <- tx_chr22[tx_id(tx_chr22) %in% coding_tx_ids]

# Define promoters as 900 bp upstream and 100 bp downstream
coding_promoters <- promoters(coding_tx_chr22, upstream = 900, downstream = 100)

# Extract promoter sequences
# getSeq() respects the strand of each promoter, so the sequence
# is represented in the same orientation as the transcript
coding_promoter_sequences <- getSeq(Hsapiens, coding_promoters)

# Determine whether each promoter contains TATAAA
has_tataaa <- vcountPattern("TATAAA", coding_promoter_sequences) > 0

# Count promoters containing at least one TATAAA
(tata_promoters <- sum(has_tataaa)) # 193

```

# How many bases on chr22 are part of more than one promoter of a coding sequence?
## Promoter: 900 bp upstream and 100 bp downstream of the transcription start site
## Ignore strand when determining overlapping promoter regions

```{r}

# Get all transcripts on chr22
tx_chr22 <- transcripts(txdb, filter = list(tx_chrom = "chr22"))

# Determine which transcripts have a coding sequence
tx_lengths <- transcriptLengths(txdb, with.cds_len = TRUE)

coding_tx_ids <- tx_lengths$tx_id[tx_lengths$cds_len > 0]

# Keep only coding transcripts
coding_tx_chr22 <- tx_chr22[tx_id(tx_chr22) %in% coding_tx_ids]

# Define 900 bp upstream / 100 bp downstream promoters
coding_promoters <- promoters(coding_tx_chr22, upstream = 900, downstream = 100)

# Ignore strand for overlap calculation
strand(coding_promoters) <- "*"

# Calculate coverage of all coding promoters
promoter_coverage <- coverage(coding_promoters)

# Count bases covered by more than one promoter
(bases_multiple_promoters <- sum(promoter_coverage > 1)) # 328099

```
