---
title: "4 Bioconductor Data Access and Analysis"
author: "Olivia Rippee"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Yeast RNAseq Data

```{r yeast RNAseq data}

library(yeastRNASeq)
fastqFilePath <- system.file("reads", "wt_1_f.fastq.gz", package = "yeastRNASeq")

```

# What fraction of reads in this file has an A nucleotide in the 5th base of the read?

```{r}

# Read the FASTQ file
x <- readLines(gzfile(fastqFilePath))

# Sequence lines are lines 2, 6, 10, 14, ...
seqs <- x[seq(2, length(x), by = 4)]

# Get the 5th base of every read
fifth_base <- substr(seqs, 5, 5)

# Fraction that is A
mean(fifth_base == "A") # 0.363841

```

# What is the average numeric quality value of the 5th base of these reads?

```{r}

library(ShortRead)

fastqFilePath <- system.file("reads", "wt_1_f.fastq.gz", package = "yeastRNASeq")

reads <- readFastq(fastqFilePath)

# Average quality score of the 5th base
mean(as(quality(reads), "matrix")[, 5]) # 28.93346

```

# leeBam Data

```{r leeBam data}

library(leeBamViews)
bamFilePath <- system.file("bam", "isowt5_13e.bam", package="leeBamViews")

```

# In the interval from 800,000 to 801,000 on yeast Chr3, how many reads are duplicated by position?

```{r}

library(Rsamtools)
library(GenomicRanges)

bamFile <- BamFile(bamFilePath)

gr <- GRanges(seqnames = "Scchr13",
  ranges = IRanges(start = 800000, end = 801000))

params <- ScanBamParam(which = gr, what = scanBamWhat())

aln <- scanBam(bamFile, param = params)

# Count reads starting at each position
pos_counts <- table(aln[[1]]$pos)

# Number of reads that have at least one duplicate
sum(pos_counts[pos_counts >= 2]) # 129

```

```{r leeBam file paths}

# File paths
bpaths <- list.files(system.file("bam", package="leeBamViews"), pattern = "bam$", full=TRUE)

```

# What is the average number of reads across the 8 samples falling in the interval Scchr13:807762-808068?

```{r}

region <- GRanges(seqnames = "Scchr13",
  ranges = IRanges(807762, 808068))

counts <- sapply(bpaths, function(bam) {
  param <- ScanBamParam(which = region, what = "qname")
  reads <- scanBam(bam, param = param)[[1]]
  length(reads$qname)})

counts
mean(counts) # 90.25

```

# In the lecture on the oligo package an ExpressionSet with 18 samples is constructed, representing normalized data from an Affymetrix gene expression microarray. The samples are divided into two groups given by the group variable. What is the average expression across samples in the control group for the “8149273” probeset?

```{r}

library(oligo)
library(GEOquery)

# Download the GEO data
getGEOSuppFiles("GSE38792")

# Extract CEL files
untar("GSE38792/GSE38792_RAW.tar", exdir = "GSE38792/CEL")

# Get CEL file paths
celfiles <- list.files("GSE38792/CEL", full.names = TRUE)

# Read CEL files
rawData <- read.celfiles(celfiles)

# Clean sample information
filename <- sampleNames(rawData)

pData(rawData)$filename <- filename

sampleNames <- sub(".*_", "", filename)
sampleNames <- sub(".CEL.gz$", "", sampleNames)

sampleNames(rawData) <- sampleNames

pData(rawData)$group <- ifelse(
  grepl("^OSA", sampleNames(rawData)), "OSA", "Control")

# RMA Normalization
normData <- rma(rawData)

# Check groups
pData(normData)$group

# Find the probeset by its (character) identifier
exprs(normData)["8149273", ]

# Calculate the average expression in the Control group
control <- pData(normData)$group == "Control"

mean(exprs(normData)["8149273", control]) # 7.02183

```

# Use the limma package to fit a two group comparison between the control group and the OSA group, and borrow strength across the genes using eBayes(). Include all 18 samples in the model fit. What is the absolute value of the log foldchange (logFC) of the gene with the lowest P.value?

```{r}

library(limma)

# Get expression matrix
expr <- exprs(normData)

# Create the group factor
group <- factor(pData(normData)$group, levels = c("Control", "OSA"))

# Design matrix
design <- model.matrix(~ group)

# Fit the model
fit <- lmFit(expr, design)

# Empirical Bayes moderation
fit <- eBayes(fit)

# Get results for the OSA vs Control comparison
results <- topTable(fit, coef = "groupOSA", number = Inf, sort.by = "P")

# View gene with smallest P-value
results[1, ]

# Absolute value of logFC
abs(results$logFC[1]) # 0.7126484

```

# How many genes are differentially expressed between the two groups at an adj.P.value cutoff of 0.05?

```{r}

results <- topTable(fit, coef = "groupOSA", number = Inf, adjust.method = "BH")

sum(results$adj.P.Val < 0.05) # 0

```

# An example 450k dataset is contained in the minfiData package. This dataset contains 6 samples; 3 cancer and 3 normals. Cancer has been shown to be globally hypo-methylated (less methylated) compared to normal tissue of the same kind. Take the RGsetEx dataset inthis package and preprocess it with the preprocessFunnorm function. For each sample, compute the average Beta value (percent methylation) across so-called OpenSea loci. What is the mean difference in beta values between the 3 normal samples and the 3 cancer samples, across OpenSea CpGs?

```{r}

library(minfi)
library(minfiData)

# Beta values after functional normalization
Beta_value <- getBeta(RGsetEX_normalized)

# OpenSea CpGs
openSea <- getIslandStatus(RGsetEX_normalized) == "OpenSea"
Beta_Value_openSea <- Beta_value[openSea, ]

# Identify samples by their actual status
normal <- pData(RGsetEx)$status == "normal"
cancer <- pData(RGsetEx)$status == "cancer"

# Average Beta values for each group
normal_mean <- mean(Beta_Value_openSea[, normal], na.rm = TRUE)
cancer_mean <- mean(Beta_Value_openSea[, cancer], na.rm = TRUE)

# Difference
normal_mean - cancer_mean # 0.08863657

```

# The Caco2 cell line is a colon cancer cell line profiled by ENCODE. Obtain the narrowPeak DNase hyper sensitive sites computed by the analysis working group (AWG). How many of these DNase hypersensitive sites contain one or more CpGs on the 450k array?

```{r}

library(AnnotationHub)

ah <- AnnotationHub()
ah <- subset(ah, species == "Homo sapiens")

# Get Caco2 AWG DNase hypersensitive sites
ah_Caco2 <- query(ah, c("Caco2", "Awg", "Dna"))
ah_Caco2 <- ah_Caco2[["AH22442"]]

# 450k CpG locations
CpG_450K <- granges(RGsetEX_normalized)

# Find 450k CpGs contained within DNase hypersensitive sites
hits <- findOverlaps(CpG_450K, ah_Caco2, type = "within")

# Count unique DNase sites containing at least one 450k CpG
length(unique(subjectHits(hits))) # 40151

```

# The zebrafishRNASeq package contains summarized data from an RNA-seq experiment in zebrafish in the form of a data.frame called zfGenes. The experiment compared 3 control samples to 3 treatment samples. Each row is a transcript; the data.frame contains 92 rows with spikein transcripts; these have a rowname starting with “ERCC”. Exclude these rows from the analysis.
## Use DESeq2 to perform a differential expression analysis between control and treatment. Do not discard (filter) genes and use the padjstart verbatim, padj, end verbatim results output as the p-value. How many features are differentially expressed between control and treatment (ie. padj <= 0.05)?

```{r}

library(zebrafishRNASeq)
library(DESeq2)

data(zfGenes)

# Remove ERCC spike-in transcripts
zfGenes_noERCC <- zfGenes[!grepl("^ERCC", rownames(zfGenes)), ]

# Define sample groups
group <- factor(c("control", "control", "control", "treatment", "treatment", "treatment"))

# Create DESeq2 dataset
dds <- DESeqDataSetFromMatrix(countData = zfGenes_noERCC,
  colData = data.frame(group), design = ~ group)

# Differential expression analysis
dds <- DESeq(dds)

# Get results
DEgenes <- results(dds)

# Number with adjusted P-value <= 0.05
sum(DEgenes$padj <= 0.05, na.rm = TRUE) # 116

# Volcano plot
plot(x = DEgenes$log2FoldChange, y = -log10(DEgenes$padj),
  pch = 20,
  col = ifelse(DEgenes$padj <= 0.05, "red", "gray"),
  xlab = "Log2 Fold Change",
  ylab = "-Log10 Adjusted P-value",
  main = "Control vs Treatment")

# Add significance cutoff
abline(h = -log10(0.05), col = "blue", lty = 2)

# Label the 10 most significant genes
sig <- which(DEgenes$padj <= 0.05)
top10 <- sig[order(DEgenes$padj[sig])[1:min(10, length(sig))]]

text(DEgenes$log2FoldChange[top10], 
     -log10(DEgenes$padj[top10]),
  labels = rownames(DEgenes)[top10],
  pos = 3,
  cex = 0.7)

```




