---
title: "3 R Data Structures and Bioconductor"
author: "Olivia Rippee"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r ALL data}
library(ALL)

# Load the ALL dataset
data(ALL)
```

# What is the mean expression across all features for sample 5 in the ALL dataset (from the ALL package)?

```{r}

# Calculate the mean expression for sample 5
mean(exprs(ALL)[, 5]) # 5.629627

```

# How many probesets (features) are annotated with more than one Ensembl gene id?

```{r}

library(biomaRt)

# Connect to Ensembl 75 (February 2014 archive)
mart <- useMart(host = "feb2014.archive.ensembl.org", 
                biomart = "ENSEMBL_MART_ENSEMBL")

# Select the human Ensembl dataset
mart <- useDataset(dataset = "hsapiens_gene_ensembl", mart = mart)

# Get the probeset IDs from the ALL dataset
probes <- featureNames(ALL)

# Annotate probesets with Ensembl gene IDs
annotation <- getBM(
  attributes = c("affy_hg_u95av2", "ensembl_gene_id"),
  filters = "affy_hg_u95av2",
  values = probes,
  mart = mart)

# Count how many Ensembl genes are associated with each probeset
genes_per_probe <- table(annotation$affy_hg_u95av2)

# Count probesets associated with more than one Ensembl gene
sum(genes_per_probe > 1) # 1045

```

# How many probesets (Affymetrix IDs) are annotated with one or more genes on the autosomes (chromosomes 1 to 22)?

```{r}

# Get chromosome information along with the Affymetrix ID
annotation <- getBM(
  attributes = c("affy_hg_u95av2", "ensembl_gene_id", "chromosome_name"),
  filters = "affy_hg_u95av2",
  values = probes,
  mart = mart)

# Keep genes located on chromosomes 1–22
autosomal <- annotation[annotation$chromosome_name %in% as.character(1:22), ]

# Count unique Affymetrix probesets
length(unique(autosomal$affy_hg_u95av2)) # 11016

```

# Use the MsetEx dataset from the minfiData package. What is the mean value of the Methylation channel across the features for sample # “5723646052_R04C01”?

```{r}

library(minfi)
library(minfiData)

# Load the MsetEx dataset
data(MsetEx)

# Extract the Methylation channel
meth <- getMeth(MsetEx)

# Calculate the mean across features for the specified sample
mean(meth[, "5723646052_R04C01"]) # 7228.277

```

# Access the processed data from NCBI GEO Accession number GSE788. What is the mean expression level of sample GSM9024?

```{r}

library(GEOquery)

# Download the processed GEO dataset
gse <- getGEO("GSE788", GSEMatrix = TRUE)

# Extract the expression data
expr <- exprs(gse[[1]])

# Calculate mean expression for GSM9024
mean(expr[, "GSM9024"]) # 756.432

```

# We are using the airway dataset from the airway package. What is the average of the average length across the samples in the experiment?

```{r airway data}

library(airway)
data(airway)

```

# What is the average of the average length across the samples in the experiment?

```{r}

# View the sample metadata
colData(airway)

# Extract average length for each sample
airway$avgLength

# Calculate the average of the average lengths
mean(airway$avgLength) # 113.75

```

# What is the number of Ensembl genes which have a count of 1 read or more in sample SRR1039512?

```{r}

# Extract counts for sample SRR1039512
counts <- assay(airway)

sample_counts <- counts[, "SRR1039512"]

# Count Ensembl genes with at least 1 read
sum(sample_counts >= 1) # 25699

```

# How many of these features overlaps with transcripts on the autosomes (chromosomes 1-22) as represented by the TxDb.Hsapiens.UCSC.hg19.knownGene package?

```{r}

library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(GenomicFeatures)
library(GenomicRanges)

# Load the hg19 knownGene annotation
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene

# Get exons grouped by transcript
exons <- exonsBy(txdb, by = "tx")

# Keep only exons on autosomes (chromosomes 1-22)
exons_autosomes <- exons[seqnames(exons) %in% paste0("chr", 1:22)]

# Combine overlapping exons into non-overlapping regions
exons_autosomes <- reduce(unlist(exons_autosomes))

# Get genomic ranges for airway features
airway_features <- rowRanges(airway)

# Keep only airway features on chromosomes 1-22
airway_autosomes <- keepSeqlevels(
  airway_features,
  as.character(1:22),
  pruning.mode = "coarse")

# Change chromosome names from 1, 2, ... to chr1, chr2, ...
seqlevels(airway_autosomes) <- paste0("chr", seqlevels(airway_autosomes))

# Find overlaps between airway features and actual exons
hits <- findOverlaps(airway_autosomes, exons_autosomes)

# Count unique airway features with at least one overlap
length(unique(queryHits(hits))) # 26276

```

# For sample SRR1039508, how big a percentage (between 0 and 1) of the total reads in the airway dataset for that sample, are part of a feature which overlaps an autosomal TxDb.Hsapiens.UCSC.hg19.knownGene transcript?

```{r}

# Counts for sample SRR1039508
sample_counts <- assay(airway)[, "SRR1039508"]

# Features overlapping autosomal TxDb transcripts
overlapping_features <- unique(queryHits(hits))

# Convert the feature indices back to the original airway row indices
overlapping_names <- names(airway_autosomes)[overlapping_features]

overlapping_indices <- match(overlapping_names, rownames(airway))

# Calculate reads in overlapping features
overlapping_reads <- sum(sample_counts[overlapping_indices], na.rm = TRUE)

# Calculate total reads
total_reads <- sum(sample_counts)

# Calculate proportion
(percentage <- overlapping_reads / total_reads) # 0.9004193

```

# What is the median number of counts per feature (for sample SRR1039508) containing a H3K4me narrowPeak in their promoter (only features which overlap autosomal transcripts from TxDb.Hsapiens.UCSC.hg19.knownGene are considered)?

```{r}

library(AnnotationHub)
library(GenomicRanges)

# Get E096 H3K4me3 narrowPeak data
ah <- AnnotationHub()
qah_h1 <- query(ah, c("E096", "H3K4me3"))
h1 <- qah_h1[["AH30596"]]

# Keep autosomes
autosome <- paste0("chr", 1:22)

h1 <- keepSeqlevels(h1, autosome, pruning.mode = "coarse")

# Get airway feature ranges that overlap autosomal TxDb transcripts
t <- rowRanges(airway)[overlapping_indices]

# Convert airway chromosome names to UCSC style
old <- seqlevels(t)
new <- paste0("chr", old)
names(new) <- old
t <- renameSeqlevels(t, new)

# Keep autosomes
t <- keepSeqlevels(t, autosome, pruning.mode = "coarse")

# Standard 2.2 kb promoter
p <- promoters(t)

# Find promoters containing H3K4me3 peaks
ov <- subsetByOverlaps(p, h1)

# Get the corresponding original airway feature indices
marked <- overlapping_indices[match(names(ov), names(t))]

# Counts for SRR1039508
marked_counts <- sample_counts[marked]

# Median count
median(marked_counts, na.rm = TRUE) # 170

```
